---
title: "Final Project 2: Gene Expression ML on TCGA-BRCA"
author: "Sihang Jiang"
date: "2026-03-26"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(550)
source("../code/00_prepare_analysis_data.R")
```

# Introduction

Breast cancer is a heterogeneous disease with substantial variation in gene expression patterns across patients.
High-dimensional transcriptomic data provide an opportunity to identify latent sample structure and to explore potential molecular subgroups using machine learning methods.

The goal of this project is to analyze TCGA-BRCA gene expression data using both unsupervised and supervised learning approaches.
Specifically, this analysis uses principal component analysis (PCA) for dimension reduction, k-means clustering for subgroup identification, and LASSO logistic regression for subgroup prediction and feature selection.

## Data

The data come from the TCGA breast cancer (BRCA) cohort and consist of normalized RNA-seq gene expression measurements.
The analysis begins with a gene-by-sample expression matrix obtained from the curatedTCGAData package.

```{r}
dim(expr)
```


The full dataset contains gene expression values for a large number of genes across breast cancer samples.
To reduce noise and focus on informative features, only the most variable genes are retained for downstream analysis.

## Methods
Variance filtering

Because many genes show limited variation across samples, genes with the highest variance were selected before model fitting.
In this project, the top 500 most variable genes were used to construct the analysis matrix.


```{r}
length(top_genes)
dim(X)
```

Here, X is the sample-by-gene matrix used for PCA, clustering, and classification.

Principal component analysis

Principal component analysis was applied to the filtered expression matrix in order to summarize the major axes of variation in the data.
The first few principal components were then used as lower-dimensional representations of the samples.

## K-means clustering

K-means clustering with 𝑘=2 was applied to the first five principal components.
This produced two clusters, which were treated as intrinsic gene-expression subgroups for the remainder of the analysis.

table(y)
##LASSO logistic regression

After cluster membership was defined, a LASSO logistic regression model was fit using the filtered gene expression matrix as predictors and cluster label as the binary outcome.
The purpose of this step was both prediction and variable selection.
LASSO imposes an L1 penalty, encouraging sparse solutions and identifying a smaller subset of genes that best distinguish the two clusters.

## Results
### Figure 1: PCA visualization

```{r figure1, echo=FALSE}
source("../code/02_make_figure.R", local = knitr::knit_global())
```

Figure 1 displays the samples projected onto the first two principal components, with points colored by the cluster labels from k-means clustering.
The plot suggests that the two clusters occupy partially distinct regions in the reduced-dimensional space, indicating meaningful heterogeneity in the underlying gene expression profiles.

### Table 1: Genes selected by LASSO

```{r table1, echo=FALSE}
source("../code/03_make_table.R", local = knitr::knit_global())
table1_gt
```
```

Table 1 reports genes with non-zero coefficients selected by the LASSO model.
These genes represent the strongest contributors to separation between the two inferred expression-based subgroups.

## Discussion

This project applied a machine learning workflow to TCGA-BRCA gene expression data.
After filtering to highly variable genes, PCA was used to reduce dimensionality and reveal major patterns of variation.
K-means clustering identified two sample subgroups, and LASSO logistic regression was then used to predict cluster membership and select important genes.

Overall, the results show that high-dimensional breast cancer gene expression data contain structured variation that can be summarized and modeled effectively using standard machine learning methods.
The PCA plot suggests separation between clusters, while the LASSO model identifies a sparse set of potentially informative genes.

Several limitations should be noted.
First, the clusters were defined by an unsupervised algorithm and may not correspond directly to known biological subtypes.
Second, the analysis used only one filtering strategy and one clustering specification.
Future work could compare alternative clustering methods, incorporate known subtype labels, and evaluate model stability using resampling or external validation.

## Conclusion

In summary, this analysis demonstrates a reproducible workflow for exploring TCGA-BRCA transcriptomic data using dimension reduction, clustering, and penalized regression.
The project highlights how unsupervised and supervised learning can be combined to identify structure in high-dimensional genomic data and to select genes associated with that structure.